---
title: "suwo"
subtitle: Access nature media from online repositories
pagetitle: Access nature media from online repositories
author: 
- Jorge Elizondo
- <a href="https://marce10.github.io/">Marcelo Araya-Salas, PhD</a>
date:  "r Sys.Date()"
output:
  rmarkdown::html_document:
    self_contained: yes
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
vignette: >
   %\VignetteIndexEntry{1. Package overview}
   %\usepackage[utf8]{inputenc}
   %\VignetteEncoding{UTF-8}
   %\VignetteDepends{Rraven}
   %\VignetteDepends{viridis}
   %\VignetteDepends{suwo}
   %\VignetteDepends{kableExtra}
   %\VignetteEngine{knitr::rmarkdown}     
editor_options: 
  chunk_output_type: console
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

&nbsp;

```{r setup, eval = TRUE, echo = FALSE, message=FALSE}

library(knitr)
library(htmlwidgets)
library(suwo)

# Create custom printing method
.print_df <- function(x, highlight = NULL, ...) {
  kbl <- kableExtra::kable(
    head(as.data.frame(x)),
    align = "c",
    row.names = FALSE,
    format = "html",
    escape = FALSE
  )
  
  if (!is.null(highlight))
    kbl <- column_spec(kbl, column = which(names(x) %in% highlight), background = "#ccebff", bold = TRUE)
  
  kbl <- kableExtra::kable_styling(kbl, bootstrap_options = "striped", font_size = 14)
  kbl <- kableExtra::scroll_box(kbl, width = "100%", height = "300px")
  
  asis_output(kbl)
}

# Register custom data frame print method
registerS3method("knit_print", "data.frame", .print_df)

# Global chunk options
knitr::opts_chunk$set(
  fig.width = 5, 
  fig.height = 3.5,
  dpi = 70,
  comment = "",
  out.width = "80%",
  fig.align = "center",
  message = TRUE,
  warning = TRUE
)

options(width = 100, max.print = 100)


```


The suwo package is designed to simplify the retrieval of observational and media data for species across multiple online biodiversity databases, including GBIF, iNaturalist, Macaulay Library, Wikiaves, Xeno-Canto, and Observation.org. This vignette provides an overview of the package’s core querying functions and demonstrates how to work with species metadata, especially for projects involving bioacoustics or species distribution modeling.

Some queries may not return results due to missing links or removed records in the consulted databases.

Installation

Install from CRAN:

```{r, eval = FALSE}

# From CRAN would be
install.packages("suwo")

#load package
library(suwo)


```

Or install the latest development version from GitHub:

```{r, eval = FALSE}

# install package
remotes::install_github("maRce10/suwo")

#load packages
library(suwo)

```

# Query Functions Overview

The following table summarizes the available suwo query functions and the types of metadata they retrieve:

```{r query_summary_table, echo=FALSE, results='asis', message=FALSE}

library(knitr)
library(kableExtra)

query_summary <- data.frame(
  `Function` = c("query_gbif", "query_inaturalist", "query_macaulay",
                 "query_observation", "query_wikiaves", "query_xenocanto"),
  `Repository` = c("GBIF", "iNaturalist", "Macaulay Library",
                   "Observation.org", "Wikiaves", "Xeno-Canto"),
  `URL link` = c(
    "https://www.gbif.org/",
    "https://www.inaturalist.org/",
    "https://www.macaulaylibrary.org/",
    "https://observation.org/",
    "https://www.wikiaves.com.br/",
    "https://www.xeno-canto.org/"
  ),
  `Requires Token` = c("No", "No", "No", "Yes", "No", "No"),
  `File Types` = c("sound, image, moving image, interactive resource",
                   "sound, image", "audio, photo, video",
                   "sound, image", "sound, image", "sound"),
  `Taxonomic Range` = c("species", "species", "species, subspecies",
                        "species", "species", "species, genus, family"),
  `Max Results` = rep("not specified", 6),
  `Geographic Coverage` = c("worldwide", "worldwide", "worldwide",
                            "worldwide", "Brazil", "worldwide"),
  `Additional Data` = c(
    "key, datasetKey, publishingOrgKey, installationKey, hostingOrganizationKey, publishingCountry, protocol, lastCrawled, lastParsed, crawlId, basisOfRecord, occurrenceStatus, taxonKey, kingdomKey, phylumKey, classKey, orderKey, familyKey, genusKey, speciesKey, acceptedTaxonKey, scientificName, acceptedScientificName, kingdom, phylum, order, family, genus, species, genericName, specificEpithet, taxonRank, taxonomicStatus, iucnRedListCategory, dateIdentified, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, continent, stateProvince, year, month, day, eventDate, modified, lastInterpreted, references, license, isInCluster, datasetName, recordedBy, identifiedBy, geodeticDatum, class, countryCode, country, rightsHolder, identifier, http...unknown.org.nick, verbatimEventDate, verbatimLocality, collectionCode, gbifID, occurrenceID, taxonID, catalogNumber, institutionCode, eventTime, occurrenceRemarks, http...unknown.org.captive, identificationID, identificationRemarks, distanceFromCentroidInMeters, informationWithheld, sex, lifeStage, preparations, nomenclaturalCode, dynamicProperties, locality, vernacularName, fieldNotes, verbatimElevation, behavior, higherClassification, associatedTaxa, infraspecificEpithet, media-type, media-format, media-references, media-created, media-creator, media-publisher, media-license, media-rightsHolder, file_url, media-description, page, elevation, elevationAccuracy, organismQuantity, organismQuantityType, georeferenceProtocol, verbatimSRS, county, verbatimCoordinateSystem, type, collectionID, individualCount, samplingProtocol, scientificNameID, georeferenceRemarks, language, georeferenceSources, media-title, repository"
,
    "quality_grade, time_observed_at, taxon_geoprivacy, uuid, id, cached_votes_total, identifications_most_agree, species_guess, identifications_most_disagree, positional_accuracy, comments_count, site_id, created_time_zone, license_code, observed_time_zone, public_positional_accuracy, oauth_application_id, created_at, description, time_zone_offset, observed_on, observed_on_string, updated_at, captive, faves_count, num_identification_agreements, map_scale, uri, community_taxon_id, owners_identification_from_vision, identifications_count, obscured, num_identification_disagreements, geoprivacy, location, spam, mappable, identifications_some_agree, place_guess, file_url, attribution, page, repository"
,
    "id, scientific_name, name, group, group_name, status, rarity, photo, info_text, permalink, determination_requirements, file_url, repository"
,
    "id, scientific_name, name, group, group_name, status, rarity, photo, info_text, permalink, determination_requirements, file_url, repository"
,
    "recording ID, media type, user ID, species ID, scientific name, common name, repository ID, author, user name, date, verified condition, location, location ID, comments, likes, visualizations, url, duration and repository"
,
    "recording ID, Genus, Specific epithet, Subspecies, English name, Recordist, Country, Locality, Latitude, Longitude, Vocalization type, Audio file, License, URL, Quality, Time, Date, Sound files in .mp3"

  ),
  stringsAsFactors = FALSE, 
check.names = FALSE
)

query_summary$`URL link` <- kableExtra::cell_spec(query_summary$`URL link`, "html", link = query_summary$`URL link`, new_tab = TRUE)

query_summary[, names(query_summary) != "Additional Data"] |>
  kbl(
    caption = "Table 1: Summary of query functions.",
    format = "html",
    escape = FALSE
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  )


query_summary[,c("Function", "Additional Data")] |>
  kbl(
    caption = "Table 2: Summary of available metadata.",
    format = "html",
    escape = TRUE
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  )


```


Finding data using suwo follows a basic sequence. First a database is selected and the corresponding query function is used to search for the data needed.

<div class="alert alert-info">

<font size = "4">Search queries</font>

A search query consists of a term (species name), a data type that corresponds to the type of data the user wishes to search for. Depending on the search function, this can vary from pictures, videos, sounds. The all_data parameter determines if all the data present for the search is included or if a summary of the metadata is generated as the output of the query. Queries may me made to any of the six available query functions that connect to one repository.

</div>


# Case study: observations of the Honduran White Bat (*Ectophylla alba*)

The Honduran White Bat is a rare, visually striking species found in Central America. It constructs tents from Heliconia leaves and emits distinct ultrasonic vocalizations. Using suwo, we can easily gather observational data, photos, and sound recordings of this species from various public databases.

Let's say we'd like to analyze the geographic distribution of this species, as well
as obtain information on the appearance and sound emited by the species.

First, we can obtain data observational data and media from the query functions by
searching for information on GBIF, INaturalist and Xenocanto.

```{r, eval = TRUE, echo = TRUE}

# Get occurrence data from GBIF
ectophylla_gbif <- query_gbif(term = "Ectophylla alba", format = "image")

# Get observations from iNaturalist
ectophylla_inat <- query_inaturalist(term = "Ectophylla alba", format = "image")

# Get audio recordings from xeno-canto
ectophylla_xc <- query_xenocanto(term = "Ectophylla alba")

ectophylla_gbif
```

```{r, echo=FALSE}

# Get occurrence data from GBIF
ectophylla_gbif <- query_gbif(term = "Ectophylla alba", format = "image")

# Get observations from iNaturalist
ectophylla_inat <- query_inaturalist(term = "Ectophylla alba", format = "image")

# Get audio recordings from xeno-canto
ectophylla_xc <- query_xenocanto(term = "Ectophylla alba")

kable_styling(
  kable(head(ectophylla_gbif, 2), caption = "First 2 rows from GBIF"),
  bootstrap_options = c("striped", "hover"),
  full_width = FALSE
)

kable_styling(
  kable(head(ectophylla_inat, 2), caption = "First 2 rows from iNaturalist"),
  bootstrap_options = c("striped", "hover"),
  full_width = FALSE
)

kable_styling(
  kable(head(ectophylla_xc, 2), caption = "First 2 rows from Xeno-Canto"),
  bootstrap_options = c("striped", "hover"),
  full_width = FALSE
)

```

## Updating queries

When past searches must be checked for new information, the function update_query allows users to update previous searches from the same function to obtain new data starting from the last search. The previous search must be loaded as an rds file because it uses the attributes of the dataframe.

```{r}

#Remove one row from previous query
ectohpylla_xc <- ectophylla_xc[-1, ]
#Update the query
updated_query_result_xenocanto  <- update_query(ectophylla_xc)

```

Metadata can be used to download the available files from the data stored from the query. This can be done utilizing the function download_media. This function may take a longer time to run depending on the available resources and the amount of data that needs to be downloaded. This information can be used to run computer vision analysis on the appearance of the species or to analyze the sound structure of the echolocation emitted by the Honduran White Bat. Updating queries must be done before merging metadata (see next section) as the required attributes are lost when merging.

We'll subset the image dataset from iNaturalist to speed up the download process.

```{r, eval = TRUE}

ectophylla_gbif <- ectophylla_gbif[1:10, ]

```


```{r, eval = TRUE, echo = TRUE}

#Download media for audios from the query from Xenocanto
paths_aud <- download_media(ectophylla_xc, path = tempdir())

#Download media for images from the query from GBIF
paths_img <- download_media(ectophylla_gbif, path = tempdir())

paths_aud

paths_img
```

## Combining metadata from multiple repositories

Users may want to combine metadata from different repositories to create a comprehensive dataset. The function `merge_metadata` combines the metadata from different repositories, ensuring that the resulting dataset contains unique observations. `merge_metadata` allows to detect possible duplicated observations from the data from different repositories, which can occur when the same observation is recorded in multiple databases. This function returns a combined data frame in which only a single copy of the duplicated observations is kept.

```{r}

# Merged_Dataframe <- detect_duplicates(X = ectophylla_gbif, Y = ectophylla_inat)


```

Users may display the coordinate information of the observations from a query using the map_locations function. This will allow the user to show the locations of the observations on a map.

```{r}

# create map
map_locations(ectophylla_inat)

```

# Some notes for specific searches

The GBIF repository dataset may me consulted to filter what repositories the user would like to consult specifically. This may be viewed by downloading the data using the function check_gbif_dataset. This function downloads the csv of the complete repository list available on GBIF.

```{r}

gbif_datasets <- check_gbif_datasets(path = tempdir())

```


A query search may require a taxon code to correctly display the species observation data. This information can be consulted using the taxon_code_search function that consults the Clements database

```{r}

taxon_code_pharomachrus <- taxon_code_search(term = "Pharomachrus mocinno")

taxon_code_pharomachrus

```


<div class="alert alert-warning">

<font size = "4">Saving and sharing query outputs</font>

Query outputs contain additional metadata as dataframe attributes. These attributes are lost when saving the dataframe as a data file (e.g. CSV, TXT, XSLX, etc) To preserve this metadata, users must save their query results as RDS files using the `saveRDS()` function from base R. This allows for easy sharing and reloading of the data with all its attributes intact. Dataframes can be retrieved from this files using the `readRDS()`. 
 

</div>



<font size="4">Session information</font>

```{r session info, echo=F}

sessionInfo()

```

